---
title: "Mouse Visium Example: Detecting Repeated Spatial Patterns"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Mouse Visium Example: Detecting Repeated Spatial Patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  run_analysis: false
  show_data_prep_code: false
---

# Introduction

This is the mouse Visium version of the workflow. It follows the same steps as the main TNBC example, using continuous features and the Euclidean metric. The one difference is preprocessing: the full object is large (~48,556 cells), so we keep only a small contiguous crop of the tissue (~5,000 cells). See the main TNBC vignette for the detailed explanation of each step.

# Load libraries

```{r setup, message=FALSE, warning=FALSE}
library(BiocStyle)
library(reticulate)
```

```{r import-package, eval=exists("params") && isTRUE(params$run_analysis), message=FALSE, warning=FALSE}
anndata <- import("anndata", convert = FALSE)
repspat <- import("repspat", convert = FALSE)
```

# Prepare the Sample

The mouse object is a single tissue, so instead of selecting one sample we crop the spatial coordinates to a rectangular window centered on the tissue, sized to keep roughly 5,000 cells. Because it is a contiguous crop, local neighborhoods and density are preserved. The feature matrix is stored in a `"layer 1"` layer and a single `sample_id` label is added. Edit `target_cells` to change the crop size.

```{r crop-settings, echo=exists("params") && isTRUE(params$show_data_prep_code)}
brain_path <- "/Users/amanpreetsingh/My Computer/VSCode/repSpat/data/brain_data.h5ad"
h5ad_path  <- "../inst/extdata/mouse_visium_crop.h5ad"

target_cells <- 5000   # edit: approximate number of cells to keep
```

```{r create-h5ad-file, eval=FALSE}
adata_full <- anndata$read_h5ad(brain_path)

coords <- py_to_r(py_get_item(adata_full$obsm, "spatial"))[, 1:2, drop = FALSE]
x <- coords[, 1]
y <- coords[, 2]

# Size a square window, centered on the tissue, to capture ~target_cells.
n_total <- nrow(coords)
frac    <- min(1, target_cells / n_total)
cx      <- median(x)
cy      <- median(y)
half_x  <- (diff(range(x)) / 2) * sqrt(frac)
half_y  <- (diff(range(y)) / 2) * sqrt(frac)

in_window <- x >= (cx - half_x) & x <= (cx + half_x) &
             y >= (cy - half_y) & y <= (cy + half_y)

cat("Cells kept:", sum(in_window), "of", n_total, "\n")

py$adata_full <- adata_full
py$in_window <- r_to_py(which(in_window) - 1L)   # 0-based indices for Python
adata <- py_eval("adata_full[in_window].copy()", convert = FALSE)

# Store the feature matrix as a named layer and add a sample label.
py_set_item(adata$layers, "layer 1", adata$X)
n_cells <- as.integer(py_to_r(adata$n_obs))
py_set_item(adata$obs, "sample_id", r_to_py(rep("mouse_brain", n_cells)))

adata$write_h5ad(h5ad_path)
adata
```

# Compute Feature Distances

We select the cropped tissue and compute pairwise distances from the `"layer 1"` layer using the `"euclidean"` metric.

```{r compute-distances, eval=exists("params") && isTRUE(params$run_analysis)}
py$adata <- adata
adata <- py_eval("adata[adata.obs['sample_id'] == 'mouse_brain'].copy()", convert = FALSE)

adata <- repspat$compute_distances(
    adata,
    layer = "layer 1",
    metric = "euclidean"
)
adata$obs$sample_id$unique()
```

```{r inspect-adata-parts, eval=exists("params") && isTRUE(params$run_analysis)}
cat("Feature matrix: first 5 cells\n")
print(py_to_r(py_get_item(adata$layers, "layer 1"))[1:5, 1:11])

cat("\nSpatial coordinates: first 5 cells\n")
print(head(py_to_r(py_get_item(adata$obsm, "spatial")), 5))

cat("\nDistance matrix: first 5 cells x first 5 cells\n")
print(py_to_r(py_get_item(adata$obsp, "repspat_distances"))[1:5, 1:5])
```

# Select Clustering Parameters

We test combinations of neighborhood size and cluster count and score each with the spatial silhouette score.

```{r silhouette-analysis, eval=exists("params") && isTRUE(params$run_analysis)}
data <- repspat$spatial_silhouette_analysis(
    adata = adata,
    n_neighbors_list = list(as.integer(6), as.integer(8)),
    n_clusters_range = as.integer(4:8)
)
cat("\n--------------------\n\n")
py_get_item(data$uns, "silhouette_scores")
```

For this crop the scores are highest at a small number of clusters, so we proceed with `n_neighbors = 8` and `n_clusters = 7`.

# Construct Spatially Constrained Clusters

We apply constrained agglomerative hierarchical clustering (CAHC) with the chosen parameters.

```{r clustering, eval=exists("params") && isTRUE(params$run_analysis)}
adata <- repspat$spatial_constrained_hac(
    adata = adata,
    n_clusters = as.integer(7),
    n_neighs = as.integer(8)
)
cat("\n--------------------\n\n")
py_get_item(data$obs, "labels")
```

The crop is divided into 7 spatially coherent clusters, each occupying a connected part of the tissue.

# Visualize the Spatially Constrained Clusters

```{r plot-spatial-regions, eval=exists("params") && isTRUE(params$run_analysis), out.width="100%"}
plt <- reticulate::import("matplotlib.pyplot")
plot_result <- repspat$plot_spatial_clusters(
    adata = adata,
    label_key = "labels",
    figsize = reticulate::tuple(10, 7),
    point_size = 4L
)
plt$subplots_adjust(right = 0.72)
plt$show()
```

# Visualize Features Across Clusters

This shows the numeric features most enriched in each cluster compared to the rest of the tissue.

```{r plot-features, eval=exists("params") && isTRUE(params$run_analysis)}
plt <- reticulate::import("matplotlib.pyplot")
feature_plots <- repspat$plot_cluster_feature_presence(
    adata = data,
    label_key = "labels",
    top_n = as.integer(10),
    figsize = reticulate::tuple(8, 3),
    point_size = 2L,
    alpha = 1.0
)

for (fig_axes in reticulate::iterate(feature_plots$values())) {
    fig <- py_get_item(fig_axes, 0L)
    plt$figure(fig$number)
    plt$show()
}
```

# Create blocks for permutation

We group cells into blocks within each cluster so permutation happens at the block level.

```{r build-blocks, eval=exists("params") && isTRUE(params$run_analysis)}
adata <- repspat$create_blocks(
    adata = adata,
    knn = as.integer(8)
)

cat("\n--------------------\n\n")
py_get_item(data$obs, "repspat_block_id")
```

# Test Spatial Invariance Between Clusters

We compare every pair of clusters with the MMD^2 statistic, using the `"IMQ"` kernel, 200 block permutations, and Benjamini–Hochberg correction.

```{r mmd-results, eval=exists("params") && isTRUE(params$run_analysis)}
adata <- repspat$multiple_comparison(
    adata = adata,
    kernel = "IMQ",
    nperm = as.integer(200),
    adj_p = "BH"
)

cat("\n--------------------\n\n")
py_to_r(py_get_item(adata$uns, "repspat_mmd_results")$head(5L)$drop(columns = list("null_dist")))
```

Cluster pairs with a small `obs_mmd_sq` and a large `adj_p` have similar feature profiles; those with a large MMD^2 and small `adj_p` are the clearly distinct clusters.

# Visualize Repeated Spatial Patterns

Cluster pairs with `adj_p >= 0.05` are repeated spatial patterns and are drawn as a network.

```{r plot-similarity-network, eval=exists("params") && isTRUE(params$run_analysis), warning=FALSE, out.width="100%"}
similarity_matrix <- py_to_r(repspat$pairwise_results_to_matrix(
    adata,
    plot = TRUE
))

similarity_matrix
```

For this dataset, every cluster pair was significantly different (`adj_p < 0.05`), so the network has no connections. No repeated spatial patterns were detected, and there is nothing to relabel.

# Conclusion

Applied to a cropped portion of the mouse Visium sample, `repSpat` found all seven clusters to be statistically distinct, so no repeated spatial patterns were recovered here. This is itself an informative result: within this small contiguous crop, the clusters carry genuinely different feature profiles. As with the TNBC example, results are best interpreted alongside cluster size, location, and biological context rather than the test result in isolation.
